Identification and analysis of pleiotropic expression quantitative trait loci

### 2023-05-10 doc/presentation_230511_centuriday2023/presentation_centuri2023_marseille.tex
ENH
DOC centuri

\newcommand*{\presentationgoldparispath}{../presentation_230120_gold2022_paris/fig/}%
\newcommand*{\mspath}{../../out/gwas417/pval_5e-08/r2_0.1/kb_1000/window_1000000/75_50}%

% Change size of footnotes
\renewcommand{\footnotesize}{\fontsize{5pt}{5pt}\selectfont}
\title{Identification and analysis of pleiotropic expression quantitative trait loci}
\subtitle{Centuri Day 2023}
\author{Aitor Gonz\'alez}
\institute{Aix Marseille Univ, INSERM, TAGC}
\date{May 11, 2023}

@@

        \begin{itemize}
            \item Differences between humans coded in 80 x 10$^6$ frequent genetic variants
            \item Genetic variants change susceptibilities to thousands of human traits
            \item For example, breast cancer:
        \end{itemize}
        %
        \begin{figure}
            \begin{center}
                \includegraphics[width=0.5\textwidth]{\presentationgoldparispath/41586_2017_Article_BFnature24284_Fig1_HTML.jpg}
            \end{center}
            \caption{Genetic variant associations with breast cancer risk (DOI:10.1038/nature24284).}
        \end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Effect of genetic variants on gene expression: eQTLs}

        \begin{itemize}
            \item Quantitative trait loci (QTL) variants associated with changes of quantitative phenotypes
            \item Expression (eQTL) are variants associated with changes of gene expression in a given tissue
            \item eQTLs provide a putative molecular mechanism of a variant
        \end{itemize}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Genetic variants with effect on trait and gene expression}

        Colocalization of eQTLs and GWAS variants

@@

        \begin{itemize}
            \item Large datasets of associations have shown that many genetic variants change simultaneously susceptibility to many diseases, for instance:
        \end{itemize}

        % full size table is table
        \begin{table}[!tbp]
            \centering
            \scriptsize

            \csvreader[separator=tab,
            tabular=crcp{0.4\textwidth},
            late after last line=\\\hline,
            head,
            table head=\\\hline \bfseries Chrom. & \bfseries Pos (hg38) & \bfseries Gene marker& \bfseries Trait categories\\\hline,
            ]{\mspath/cmpt_count_per_rsid.py/count_per_rsid_gwas_ms.tsv}{}% use head of csv as column names
                {\csvcoli\ & \csvcolii\ & \csvcolv\ & \csvcolvi}% specify your coloumns here
            \vspace{15pt}\label{tab:pleitropic_variants}
        \end{table}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Summary of previous and my work here}

        \begin{center}
            \includegraphics[width=0.6\textwidth]{fig/graphical_intro_approach.drawio.png}
        \end{center}

@@

    \section{Results} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Most pleiotroic eQTLs}

        \begin{itemize}
            \item Example pleiotropic eQTLs in different cytobands
            \item The representative gene is the eQTL gene with the highest PubMed publication count
        \end{itemize}

% full size table is table
        \begin{table}[!tbp]
            \centering
            \scriptsize

            \csvreader[separator=tab,
            tabular=crcp{0.4\textwidth},
            late after last line=\\\hline,
            head,
            table head=\\\hline \bfseries Chrom. & \bfseries Pos (hg38) & \bfseries Gene marker& \bfseries Trait categories\\\hline,
            ]{\mspath/cmpt_count_per_rsid.py/count_per_rsid_gwas_ms.tsv}{}% use head of csv as column names
                {\csvcoli\ & \csvcolii\ & \csvcolv\ & \csvcolvi}% specify your coloumns here
            \vspace{15pt}\label{tab:pleitropic_variants}
        \end{table}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Summary of my work}

        \begin{center}
            \includegraphics[width=0.5\textwidth]{fig/graphical_abstract.drawio.png}
        \end{center}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Applications / Collaborations}

        \begin{itemize}
            \item Dataset of traits, categories, tissues and cell types
            \item Annotation of genomic regions
            \item Annotation of genes
        \end{itemize}

@@

        \begin{itemize}
            \item P Paul
            \item L Lecerf (M1), P Rihet, M Michel, S Marquet, S Spicuglia
        \end{itemize}
%
        \vfill
%
        Funding
%
        \begin{itemize}
            \item Institut Cancer et Immunologie - Aix-Marseille Univ.
            \item Agence nationale de la recherche (ANR)
            \item Centre de Calcul Intensif d'Aix-Marseille
        \end{itemize}

>>>>

\newcommand*{\presentationgoldparispath}{../presentation_230120_gold2022_paris/fig/}%
\newcommand*{\mspath}{../../out/gwas417/pval_5e-08/r2_0.1/kb_1000/window_1000000/75_50}%
\newcounter{frame}[frame]
\setbeamertemplate{footline}[frame number]

% Change size of footnotes
\renewcommand{\footnotesize}{\fontsize{5pt}{5pt}\selectfont}
\title{Identification and analysis of pleiotropic expression quantitative trait loci}
\subtitle{CENTURI Scientific Day 2023}
\date{May 11, 2023 - Hexagone auditorium, Luminy campus}
\author{Aitor Gonz\'alez}
\institute{Aix Marseille Univ, INSERM, TAGC}

@@

        \begin{itemize}
            \item Trait differences in human due to 80 x 10$^6$ genetic variants
            \item Genetic variants change susceptibilities to thousands of human traits
            \item For example, breast cancer:
        \end{itemize}
        %
        \begin{figure}
            \begin{center}
                \includegraphics[width=0.5\textwidth]{\presentationgoldparispath/41586_2017_Article_BFnature24284_Fig1_HTML.jpg}
            \end{center}
            \caption{Genetic variant associations with breast cancer risk (DOI:10.1038/nature24284).}
        \end{figure}

        $\rightarrow$ What is the molecular mechanism of the trait association?

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{eQTLs: Genetic variants that change gene expression}

@@

        \begin{itemize}
            \item eQTLs useful to indicate putative molecular mechanism of a trait variants
        \end{itemize}

        $\rightarrow$ How do we annotate trait variants with eQTLs?

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Annotation of trait variants with eQTLs}

        $\rightarrow$ Colocalization of eQTLs and trait variants

@@

        \begin{itemize}
            \item Many genetic variants change simultaneously susceptibility to several diseases, for instance:
        \end{itemize}

        \begin{table}[!tbp]
            \centering
            \scriptsize
            \csvreader[separator=tab,
            tabular=crrcp{0.4\textwidth},
            late after last line=\\\hline,
            head,
            table head=\\\hline \bfseries Chrom. & \bfseries Pos (hg38) & \bfseries Ref. & \bfseries Alt. & \bfseries Trait categories\\\hline,
            ]{fig/count_per_rsid_gwas_ms.tsv}{}% use head of csv as column names
                {\csvcoli\ & \csvcolii\ & \csvcoliii\ & \csvcoliv\ & \csvcolvi}% specify your coloumns here
            \vspace{15pt}\label{tab:2}
        \end{table}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Positioning of our work here in the state of the art}

        \begin{center}
            \includegraphics[width=0.6\textwidth]{fig/graphical_intro_approach.v2.drawio.png}
        \end{center}

@@

    \section{Results and discussion} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Pleiotroic eQTLs}

        \begin{itemize}
            \item Analyzed: 138,274 variants from 293 traits and the 127 cell types/tissues
            \item Found: 476 pleiotropic eQTLs associated to $\geq$ two traits categories
            \item Important genomic regions: ABO, ALDH2, IL4, TP53, ...
        \end{itemize}

        \begin{table}[!tbp]
            \centering
            \scriptsize
            \csvreader[separator=tab,
            tabular=crrrcp{0.4\textwidth},
            late after last line=\\\hline,
            head,
            table head=\\\hline \bfseries Chrom. & \bfseries Pos (hg38) & \bfseries Ref. & \bfseries Alt. & \bfseries Gene marker& \bfseries Trait categories\\\hline,
            ]{fig/count_per_rsid_gwas_ms.tsv}{}% use head of csv as column names
                {\csvcoli\ & \csvcolii\ & \csvcoliii\ & \csvcoliv\ & \csvcolv\ & \csvcolvi}% specify your coloumns here
            \vspace{15pt}\label{tab:1}
        \end{table}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{Graphical summary}

        \begin{center}
            \includegraphics[width=0.6\textwidth]{fig/graphical_abstract.drawio.png}
        \end{center}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \begin{frame}
        \frametitle{\includegraphics[width=0.1\textwidth]{fig/opentonetwork.png} Applications of these datasets}

        \begin{itemize}
            \item Large dataset of genetic variants ...
            \item ... associated to traits and gene expression changes in 127 tissues and cell types
            \item Useful to annotate genomic regions and genes
        \end{itemize}

@@

%
        Collaborators and colleagues at TAGC, specially:
        %
        \begin{itemize}
            \item P. Paul, L. Lecerf (M1), P. Rihet, M. Michel, S. Marquet, S. Spicuglia.
        \end{itemize}
%
        \vfill
%
        Funding and resources:
%
        \begin{itemize}
            \item Institut Cancer et Immunologie - Aix-Marseille Univ.
            \item Agence nationale de la recherche (ANR)
            \item Centre de Calcul Intensif d'Aix-Marseille
        \end{itemize}
%
        \vfill
        %
        Thank you for your attention!!
